---
title: "1. Build an Environmental Lexicon"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Aim

In this notebook you will turn interpretation into a structured lexicon.

A lexicon is not just a list of words. It is a working definition of what your
group means by environmental language.

The important question is not simply:

> Which environmental words can we think of?

but:

> What role should each word play in the analysis?

```{r}
#| label: setup
#| include: false
library(tidyverse)

dir.create("data", showWarnings = FALSE)
dir.create("output", showWarnings = FALSE)

normalise_term <- function(x) {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_squish()
}
```

## Step 1: Understand The Categories

We will classify terms in two ways.

First, each term has a `type`.

- `entity`: an environmental thing or place. Examples: forest, river, garden.
- `process`: an environmental event or transformation. Examples: storm, rain, flood.
- `condition`: a quality or state. Examples: dark, green, barren.
- `exclude`: a word you discussed but chose not to use.

Second, each term has a `role`.

- `trigger`: a word used to retrieve likely environmental sentences.
- `descriptor`: a word used to analyse how an environment is described.
- `exclude`: a word that should not be used in the workflow.

This distinction matters because some words are meaningful but bad search terms.
For example, `dark` may be important in `dark forest`, but searching for every
sentence containing `dark` will also retrieve dark rooms, dark hair, and dark
secrets.

## Step 2: Create Your Lexicon

The easiest option is to work in a spreadsheet.

Open:

```text
templates/lexicon_template.xlsx
```

Fill in the `lexicon` sheet, then save your completed file as:

```text
data/student_lexicon.xlsx
```

If Excel is not convenient in Posit Cloud, you can instead complete the CSV
template and save it as `data/student_lexicon.csv`:

```text
templates/lexicon_template.csv
```

The notebook will look first for `data/student_lexicon.xlsx`, then for
`data/student_lexicon.csv`. If neither file exists, it will use the small empty
table below, which you can edit directly in R.

```{r}
if (file.exists("data/student_lexicon.xlsx")) {
  if (!requireNamespace("readxl", quietly = TRUE)) {
    stop("Install the readxl package to read .xlsx files, or save your lexicon as data/student_lexicon.csv.")
  }
  student_lexicon <- readxl::read_excel("data/student_lexicon.xlsx", sheet = "lexicon")
} else if (file.exists("data/student_lexicon.csv")) {
  student_lexicon <- readr::read_csv("data/student_lexicon.csv", show_col_types = FALSE)
} else {
  student_lexicon <- tibble::tribble(
    ~term, ~type, ~role, ~notes,
    # "forest", "entity", "trigger", "clear environmental place",
    # "river", "entity", "trigger", "clear environmental place",
    # "storm", "process", "trigger", "weather event",
    # "dark", "condition", "descriptor", "important but too broad as a trigger",
    # "green", "condition", "descriptor", "quality or colour",
    # "village", "exclude", "exclude", "borderline built place"
  )
}
```

## Step 3: Check Your Lexicon

The code below checks that your categories are written consistently.

```{r}
student_lexicon <- student_lexicon %>%
  mutate(
    term = normalise_term(term),
    type = normalise_term(type),
    role = normalise_term(role)
  ) %>%
  filter(!is.na(term), term != "") %>%
  distinct(term, .keep_all = TRUE)

valid_types <- c("entity", "process", "condition", "exclude")
valid_roles <- c("trigger", "descriptor", "exclude")

invalid_types <- student_lexicon %>%
  filter(!type %in% valid_types)

invalid_roles <- student_lexicon %>%
  filter(!role %in% valid_roles)

if (nrow(student_lexicon) == 0) {
  stop("Your lexicon is empty. Add some terms in Step 2.")
}

if (nrow(invalid_types) > 0) {
  print(invalid_types)
  stop("Some terms have an invalid type. Use entity, process, condition, or exclude.")
}

if (nrow(invalid_roles) > 0) {
  print(invalid_roles)
  stop("Some terms have an invalid role. Use trigger, descriptor, or exclude.")
}

student_lexicon
```

## Step 4: Summarise Your Decisions

These summaries help you see the shape of your operational definition.

```{r}
student_lexicon %>%
  count(type, sort = TRUE)
```

```{r}
student_lexicon %>%
  count(role, sort = TRUE)
```

## Step 5: Save Your Lexicon

This saves (and overwrites) a CSV copy so the next notebook can use it. Even if
you created your lexicon in Excel, the later notebooks will use this CSV version.

```{r}
readr::write_csv(student_lexicon, "data/student_lexicon.csv")
```

## Reflection

Write brief answers below.

1. Which terms were easy to classify?
2. Which terms were difficult or borderline?
3. Which words did you exclude, and why?
4. What might your lexicon miss?
