---
title: "2. Analyse Environmental Sentences"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Aim

In this notebook you will use your lexicon to retrieve environmental sentences
from the prepared corpus.

The key idea is:

> Trigger terms help us find possible environmental passages. Descriptor terms
> help us interpret how environments are represented.

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(tidytext)

dir.create("output", showWarnings = FALSE)

normalise_term <- function(x) {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_squish()
}

regex_escape <- function(x) {
  stringr::str_replace_all(x, "([\\\\.^$|()\\[\\]{}*+?])", "\\\\\\1")
}

standardise_lexicon_table <- function(x) {
  required_columns <- c("term", "type", "role", "notes")

  x <- x %>%
    as_tibble(.name_repair = "minimal")

  if (ncol(x) < 4) {
    stop("The lexicon must have four columns: term, type, role, notes.")
  }

  x <- x[, seq_len(4), drop = FALSE]
  names(x) <- required_columns

  if (nrow(x) == 0) {
    return(x)
  }

  x %>%
    mutate(across(everything(), as.character)) %>%
    filter(
      !(
        normalise_term(term) == "term" &
          normalise_term(type) == "type" &
          normalise_term(role) == "role" &
          normalise_term(notes) == "notes"
      )
    )
}

read_lexicon_file <- function(path) {
  if (stringr::str_ends(path, "\\.xlsx$")) {
    if (!requireNamespace("readxl", quietly = TRUE)) {
      stop("Install the readxl package to read .xlsx files, or save your lexicon as data/student_lexicon.csv.")
    }

    sheet_names <- readxl::excel_sheets(path)
    sheet_to_read <- if ("lexicon" %in% sheet_names) "lexicon" else sheet_names[1]

    raw_lexicon <- readxl::read_excel(
      path,
      sheet = sheet_to_read,
      col_names = FALSE,
      .name_repair = "minimal"
    )
  } else {
    raw_lexicon <- readr::read_csv(
      path,
      col_names = FALSE,
      comment = "#",
      show_col_types = FALSE
    )
  }

  standardise_lexicon_table(raw_lexicon)
}
```

## Step 1: Load Your Lexicon And The Corpus

The corpus has already been prepared. It contains sentence-level extracts from
the four novels.

For the main workshop, use the prepared English corpus:

```text
data/children_lit_sentences.csv
```

For a future German version of the exercise, create a sentence-level corpus with
the same columns and change `corpus_path` to point to it, for example:

```text
data/german_children_lit_sentences.csv
```

```{r}
corpus_path <- "data/children_lit_sentences.csv"

# Optional future German corpus:
# corpus_path <- "data/german_children_lit_sentences.csv"

if (file.exists("data/student_lexicon.csv")) {
  lexicon <- read_lexicon_file("data/student_lexicon.csv")
} else if (file.exists("data/student_lexicon.xlsx")) {
  lexicon <- read_lexicon_file("data/student_lexicon.xlsx")
} else {
  stop("No lexicon found. Create data/student_lexicon.csv or data/student_lexicon.xlsx in Notebook 1.")
}

lexicon <- lexicon %>%
  mutate(
    term = normalise_term(term),
    type = normalise_term(type),
    role = normalise_term(role)
  )

corpus <- readr::read_csv(corpus_path, show_col_types = FALSE)

glimpse(corpus)
```

### Optional: German Texts

The same method can work with German children's literature, but the lexicon
usually needs more variants than the English example.

Useful checks:

- include singular and plural forms when needed, such as `wald` and `waelder`
- include common compounds if they matter, such as `waldweg`, `flussufer`,
  `gartenmauer`, or `baumkrone`
- check whether your corpus keeps umlauts (`grün`, `Wälder`) or transliterates
  them (`gruen`, `Waelder`)
- keep the `type` and `role` values in English: `entity`, `process`,
  `condition`, `trigger`, `descriptor`, `exclude`

## Step 2: Inspect The Corpus

Before searching computationally, look at the data as readers. 

```{r}
corpus %>%
  count(title, author, sort = TRUE)
```
"n" here represents the number of sentences for each title and author combination.

We can also have a look at a few sentences:

```{r}
set.seed(2026)

corpus %>%
  slice_sample(n = 8) %>%
  select(sentence, title)

```

## Step 3: Use Trigger Terms To Retrieve Sentences

Here we separate retrieval from interpretation.

Trigger terms answer:

> Which sentences should we retrieve?

```{r}
trigger_terms <- lexicon %>%
  filter(role == "trigger") %>%
  pull(term) %>%
  unique()

if (length(trigger_terms) == 0) {
  stop("Your lexicon has no trigger terms. Return to Notebook 1 and add some.")
}

trigger_terms
```

The next code block turns your trigger terms into a search pattern.

```{r}
trigger_pattern <- paste0(
  "\\b(",
  paste(regex_escape(trigger_terms), collapse = "|"),
  ")\\b"
)

environmental_sentences <- corpus %>%
  mutate(sentence_lower = normalise_term(sentence)) %>%
  filter(stringr::str_detect(sentence_lower, trigger_pattern)) %>%
  mutate(
    trigger_matches = stringr::str_extract_all(sentence_lower, trigger_pattern),
    trigger_matches = purrr::map_chr(trigger_matches, ~ paste(unique(.x), collapse = "; "))
  ) %>%
  select(sentence_id, title, author, gutenberg_id, source_url, book_sentence_id, sentence, trigger_matches)

environmental_sentences %>%
  count(title, sort = TRUE)
```

## Step 4: Inspect The Retrieved Sentences

This is where interpretation returns. Ask whether the computer has applied your
definition in a useful way.

```{r}
message("Retrieved sentences: ", nrow(environmental_sentences))
message("Total corpus sentences: ", nrow(corpus))
message("Retrieval rate: ", round(nrow(environmental_sentences) / nrow(corpus) * 100, 2), "%")
```

```{r}
set.seed(2026)

environmental_sentences %>%
  slice_sample(n = min(12, nrow(environmental_sentences))) %>%
  select(title, sentence, trigger_matches)
```

Questions:

1. Which retrieved sentences are clearly environmental?
2. Which retrieved sentences are noisy or irrelevant?
3. What might your trigger list have missed?

```{r}
readr::write_csv(environmental_sentences, "output/02_environmental_sentences.csv")
```

## Step 5: Look For Descriptor-Trigger Patterns

Descriptor terms answer:

> How is the environment being represented?

The code below tokenises sentences into individual words. It then looks for
descriptor terms within the three words before a trigger term.

This is a simple method. It will not capture every meaningful phrase, but it
helps us move from isolated keywords to relational patterns.

```{r}
descriptor_terms <- lexicon %>%
  filter(role == "descriptor") %>%
  pull(term) %>%
  unique()

if (length(descriptor_terms) == 0) {
  stop("Your lexicon has no descriptor terms. Return to Notebook 1 and add some.")
}

descriptor_terms
```

```{r}
tokens <- environmental_sentences %>%
  select(sentence_id, title, author, gutenberg_id, source_url, book_sentence_id, sentence) %>%
  tidytext::unnest_tokens(
    output = "word",
    input = "sentence",
    token = "words",
    to_lower = TRUE
  ) %>%
  group_by(sentence_id) %>%
  mutate(
    token_position = row_number(),
    prev_1 = lag(word, 1),
    prev_2 = lag(word, 2),
    prev_3 = lag(word, 3)
  ) %>%
  ungroup()

descriptor_trigger_pairs <- tokens %>%
  filter(word %in% trigger_terms) %>%
  select(sentence_id, title, author, gutenberg_id, source_url, book_sentence_id,
         token_position, trigger = word, prev_1, prev_2, prev_3) %>%
  pivot_longer(
    cols = starts_with("prev_"),
    names_to = "distance",
    values_to = "descriptor"
  ) %>%
  filter(descriptor %in% descriptor_terms) %>%
  mutate(
    distance = readr::parse_number(distance),
    pair = paste(descriptor, trigger)
  ) %>%
  arrange(sentence_id, token_position, distance) %>%
  distinct(sentence_id, token_position, descriptor, trigger, .keep_all = TRUE)

pair_counts <- descriptor_trigger_pairs %>%
  count(pair, descriptor, trigger, sort = TRUE)

pair_counts
```

```{r}
readr::write_csv(descriptor_trigger_pairs, "output/03_descriptor_trigger_pairs_long.csv")
readr::write_csv(pair_counts, "output/03_descriptor_trigger_pair_counts.csv")
```

## Reflection

1. Which descriptor-trigger pairs are most interesting?
2. Are the most frequent pairs also the most meaningful?
3. What kinds of description does this method miss?
