---
title: "3. Interpret Environmental Tropes"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Aim

In this notebook you will move from computational patterns back to literary
interpretation.

The computer can extract repeated descriptor-trigger pairs. It cannot decide
what those patterns mean. That interpretive step belongs to you.

```{r}
#| label: setup
#| include: false
library(tidyverse)

dir.create("data", showWarnings = FALSE)
dir.create("output", showWarnings = FALSE)

normalise_term <- function(x) {
  x %>%
    stringr::str_to_lower() %>%
    stringr::str_squish()
}
```

## Step 1: Load The Pattern Output

This file was created in Notebook 2.

```{r}
pair_counts <- readr::read_csv(
  "output/03_descriptor_trigger_pair_counts.csv",
  show_col_types = FALSE
)

pair_counts
```

## Step 2: Create An Annotation Template

The next block saves a template. Open the CSV and add a `proposed_trope` for
each pattern you want to interpret.

Possible trope labels might include:

- threatening wilderness
- renewal
- exposure
- scarcity
- climatic intensity
- atmospheric beauty
- hidden garden
- untamed landscape

You may invent your own labels. Disagreement is not a failure. It shows where
interpretation enters the workflow.

```{r}
trope_template <- pair_counts %>%
  arrange(desc(n), pair) %>%
  mutate(
    proposed_trope = "",
    evidence_or_reason = ""
  )

readr::write_csv(trope_template, "data/trope_annotations_template.csv")

trope_template
```

Save your completed version as:

```text
data/trope_annotations.csv
```

## Step 3: Summarise Your Trope Labels

Run this section after you have created `data/trope_annotations.csv`.

```{r}
if (!file.exists("data/trope_annotations.csv")) {
  message("No completed annotation file found yet.")
  message("Create data/trope_annotations.csv from the template, then rerun this section.")
  trope_summary <- tibble()
} else {
  raw_trope_annotations <- readr::read_csv(
    "data/trope_annotations.csv",
    show_col_types = FALSE
  ) %>%
    mutate(proposed_trope = normalise_term(proposed_trope)) %>%
    filter(!is.na(proposed_trope), proposed_trope != "")

  trope_annotations <- raw_trope_annotations %>%
    select(pair, descriptor, trigger, proposed_trope, evidence_or_reason) %>%
    inner_join(
      pair_counts,
      by = c("pair", "descriptor", "trigger")
    )

  unmatched_annotations <- raw_trope_annotations %>%
    anti_join(
      pair_counts,
      by = c("pair", "descriptor", "trigger")
    )

  if (nrow(unmatched_annotations) > 0) {
    message("Some annotations did not match the current pair output and were ignored:")
    print(unmatched_annotations %>% select(pair, proposed_trope))
  }

  if (nrow(trope_annotations) == 0) {
    message("No current descriptor-trigger pairs have been annotated yet.")
    trope_summary <- tibble()
  } else {
    trope_summary <- trope_annotations %>%
      group_by(proposed_trope) %>%
      summarise(
        total_pair_tokens = sum(n),
        unique_pairs = n_distinct(pair),
        examples = paste(head(pair, 5), collapse = "; "),
        .groups = "drop"
      ) %>%
      arrange(desc(total_pair_tokens))
  }

  trope_summary
}
```

```{r}
if (nrow(trope_summary) > 0) {
  readr::write_csv(trope_summary, "output/04_trope_summary.csv")
}
```

## Step 4: Interpret The Results

Use the table above as evidence, not as an answer.

Questions:

1. Which tropes seem most prominent?
2. Which individual pairs are especially revealing?
3. Which labels feel too broad or too narrow?
4. What would change if you revised your lexicon?
5. What did the computational method make visible?
6. What did it flatten, miss, or distort?

## Closing Thought

Digital Humanities is not a movement away from interpretation. In this workflow,
interpretation happens at every stage:

```text
choosing terms -> assigning roles -> retrieving sentences -> extracting patterns -> naming tropes
```
